
---
title: "DATA 607 - Project 1: Chess Tournament Data Extraction"
author: "Muhammad Imran"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

# Part 1: Comprehensive Project Approach

In order to process unstructured or semi-structured text data into a structured format using R, one has to follow a systematic data processing pipeline. The following chart shows the pipeline's execution flow.

\vspace{-1.5em}
```{r flowchart, echo=FALSE, message=FALSE, warning=FALSE}
library(DiagrammeR)

grViz("
digraph chess_pipeline {

  # Graph styling
  graph [layout = dot, rankdir = TB]

  # Node styling
  node [shape = rectangle, style = filled, fillcolor = LightSkyBlue, fontname = Helvetica]

  # Nodes definition
  A [label = 'Raw Data Ingestion\n(tournamentinfo.txt)']
  B [label = 'Filter Decorative Noise\n(Remove header rows & --- lines)']
  C [label = 'Line Pairing\n(Split odd & even indexed lines)']
  D [label = 'Regex Feature Extraction\n(Name, State, Pts, Pre-Rating)']
  E [label = 'Opponent Matchup Parsing\n(Extract round pair IDs)']
  F [label = 'Relational Lookup & Computation\n(Calculate mean opponent pre-rating)']
  G [label = 'Export Structured Dataset\n(chess_tournament_summary.csv)']

  # Edge definitions
  A -> B -> C -> D -> E -> F -> G
}
")

```
\vspace{-1.5em}

# Detailed Step by Step Breakdown
## Step 1: Data Ingestion and Sanitization

When reading semi-structured text in R, readLines() ingests the document as a vector of character strings, where each element corresponds to one line of the text file.

```` 
raw_data <- readLines("tournamentinfo.txt")


````

The raw text file contains decorative border dividers (lines composed of dashes ---) and header labels. We clean the vector using pattern matching:


````  
# Remove border lines starting with dashes
cleaned_lines <- raw_data[!str_detect(raw_data, "^-+")]

# Remove the two column header rows at the top
cleaned_lines <- cleaned_lines[-c(1, 2)]


````

* str_detect(raw_data, "^-+"): Finds all rows starting with one or more dashes. The ! negates this condition to keep only content rows.   
* likewise [-c(1, 2)]: Drops the first two rows, which contain column labels like Pair | Player Name and USCF ID.

##  Step 2: Reconstructing Records via Dual-Vector Pairing
Each player entry spans two consecutive rows

* Row 1: Player ID, Name, Total Points, and the 7 Round results.   
* Row 2: State, USCF ID, Pre-Rating, and Post-Rating

To handle this cleanly without looping line-by-line, we separate the vector into odd and even elements:


```

line1 <- cleaned_lines[seq(1, length(cleaned_lines), by = 2)] # Rows 1, 3, 5...
line2 <- cleaned_lines[seq(2, length(cleaned_lines), by = 2)] # Rows 2, 4, 6...


```

By doing this, line1[i] and line2[i] always refer to player i.


## Step 3: Regular Expression (Regex) Extraction
We extract key attributes from line1 and line2 using specific regex patterns:




###   1. Pair ID (from line1)
 

````  
pair_id <- as.numeric(str_trim(str_extract(line1, "^\\s*\\d+")))


````

 Matches leading spaces (`^\s*`) followed by digits (`\d+`) at the beginning of the string.   

### 2. Player Name (from line1)
````  
player_name <- str_trim(str_extract(line1, "(?<=\\|)[^|]+(?=\\|)"))


````
* Lookbehind assertion checking for a pipe (|) before the name. 
* Matches one or more characters that are not a pipe. 
* Lookahead assertion checking for a pipe (|) following the name.  
* Removes leading and trailing padding spaces.   

### 3. Player State (from line2)

````
state <- str_trim(str_extract(line2, "^\\s*[A-Z]{2}"))

````

Matches leading spaces followed by exactly two uppercase state letters (e.g., ON, MI, OH).

### 4. Total Points (from line1)
 
 ````
 total_pts <- as.numeric(str_trim(str_extract(line1, "(?<=\\|)\\d+\\.\\d+"))
 
 ````
 
 The argument, finds decimal point values (e.g., 6.0, 5.5) following a pipe delimiter.
 
### 5. Pre-Tournament Rating (from line2)
 
 ````
 pre_rating <- as.numeric(str_extract(str_extract(line2, "R:\\s*\\d+"), "\\d+"))
 
 ````
* Edge Case Handling: Some ratings include provisional game counts (e.g., 1641P17 or 1220P13).   * * * First pass: str_extract(line2, "R:\\s*\\d+") grabs R: followed by the numeric digits, stopping before any P letter.   
* Second pass: str_extract(..., "\\d+") strips out R: to leave only pure integer digits.
 
 
## Step 4: Extracting Opponents & Calculating Means

### 1. Extracting Opponent Pair IDs

In line1, character position 48 onward contains the round-by-round results:
 
````
opponents_list <- str_extract_all(str_sub(line1, start = 48), "\\b\\d+\\b")

````

* str_sub(line1, start = 48): Targets the 7 round result columns.   
* Extracts word-bounded digits (e.g., in W  39, it extracts 39; in D  12, it extracts 12).
* Letters representing non-played rounds (e.g., H for half-point bye, U for unplayed, B for bye) are automatically ignored because they don't contain opponent pair IDs.

### 2. Vectorized Lookup & Mean Calculation

We define a helper function to calculate the mean opponent rating for a single player:

````
calc_avg_opp_rating <- function(opp_ids, rating_lookup) {
  ids <- as.numeric(opp_ids)
  ratings <- rating_lookup[ids]
  return(round(mean(ratings, na.rm = TRUE)))
}

````
How this lookup works step-by-step for Gary Hua (Player 1):   

a.  Extracted Opponent IDs:      c(39, 21, 18, 14, 7, 12, 4).  
b.  Lookup Ratings:              rating_lookup[c(39, 21, 18, 14, 7, 12, 4)] maps directly to c(1436,                                 1563, 1600, 1610, 1649, 1663, 1716).   
c.  Compute Mean:                $$\text{Mean} = \frac{1436 + 1563 + 1600 + 1610 + 1649 + 1663 + 1716}{7} = 1605.285...$$




d.  Round Result:               

We then apply this helper function across all players using sapply():

````
avg_opp_rating <- sapply(opponents_list, calc_avg_opp_rating, rating_lookup = pre_rating) 

````


##  Step 5: Assembling and Exporting the Final Output

Finally, we combine all extracted components into a tidy tibble dataframe and export it to a standard CSV file:
````
final_df <- tibble(
  `Player's Name` = player_name,
  `Player's State` = state,
  `Total Number of Points` = total_pts,
  `Player's Pre-Rating` = pre_rating,
  `Average Pre Chess Rating of Opponents` = avg_opp_rating
)

# Save dataframe to disk
write_csv(final_df, "chess_tournament_summary.csv")

````




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(stringr)

# Read raw file lines
raw_url <- "[https://raw.githubusercontent.com/datasets/tournamentinfo/main/tournamentinfo.txt](https://raw.githubusercontent.com/datasets/tournamentinfo/main/tournamentinfo.txt)"

# If local file exists, read locally; otherwise read from URL
if (file.exists("tournamentinfo.txt")) {
  raw_data <- readLines("tournamentinfo.txt")
} else {
  raw_data <- readLines(raw_url)
}

# Remove horizontal dash lines
cleaned_lines <- raw_data[!str_detect(raw_data, "^-+")]

# Remove header rows (first two remaining lines)
cleaned_lines <- cleaned_lines[-c(1, 2)]

head(cleaned_lines, 6)

line1 <- cleaned_lines[seq(1, length(cleaned_lines), by = 2)]
line2 <- cleaned_lines[seq(2, length(cleaned_lines), by = 2)]



# Extract Pair ID
pair_id <- as.numeric(str_trim(str_extract(line1, "^\\s*\\d+")))

# Extract Name
player_name <- str_trim(str_extract(line1, "(?<=\\|)[^|]+(?=\\|)"))

# Extract State
state <- str_trim(str_extract(line2, "^\\s*[A-Z]{2}"))

# Extract Total Points
total_pts <- as.numeric(str_trim(str_extract(line1, "(?<=\\|)\\d+\\.\\d+")))

# Extract Pre-Rating (handles provisional ratings like '1641P17')
pre_rating <- as.numeric(str_extract(str_extract(line2, "R:\\s*\\d+"), "\\d+"))

# Combine into initial data frame
players_df <- tibble(
  PairID = pair_id,
  Name = player_name,
  State = state,
  TotalPoints = total_pts,
  PreRating = pre_rating
)

head(players_df)


# Extract opponent pair IDs from round columns in line 1
opponents_list <- str_extract_all(str_sub(line1, start = 48), "\\b\\d+\\b")

# Helper function to compute average opponent pre-rating
calc_avg_opp_rating <- function(opp_ids, rating_lookup) {
  ids <- as.numeric(opp_ids)
  ratings <- rating_lookup[ids]
  return(round(mean(ratings, na.rm = TRUE)))
}

# Compute average for all players
avg_opp_rating <- sapply(opponents_list, calc_avg_opp_rating, rating_lookup = players_df$PreRating)

# Add to final dataset
final_df <- players_df %>%
  mutate(AvgOpponentPreRating = avg_opp_rating) %>%
  select(Name, State, TotalPoints, PreRating, AvgOpponentPreRating)

head(final_df, 10)


final_df[1, ]

write_csv(final_df, "chess_tournament_summary.csv")